---
title: "Enriched Pathways"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Enriched Pathways}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

The **Enriched pathways** tab displays functional enrichment results from
methods such as Enrichr, GSVA, or any tool producing gene set enrichment
tables. It appears _conditionally_ in the sidebar — only when the loaded
`.crb` file contains pathway enrichment data.

Results are shown in a paginated, searchable DT table with automatic
formatting: percentages as progress bars, p-values as colour tiles, and
log fold-change values on a diverging colour scale.

# Quick start

```{r, eval = FALSE}
library(CerebroNexus)
launchCerebroV1.4()
```

1. Launch CerebroNexus and load a `.crb` file with enrichment results
2. If enrichment data is present, **Enriched pathways** appears in the sidebar
3. Choose an enrichment method (e.g., `cerebro_seurat_enrichr`)
4. Choose a grouping variable (e.g., `seurat_clusters`)
5. Browse, search, sort, and download the results

# Table formatting

Columns are auto-detected and formatted:

- **Percentages**: colour-filled progress bars
- **P-values**: red (significant) to white
- **Log fold-change**: blue (negative) to red (positive)
- **Long text**: hidden behind a toggle to keep the table compact

# Data export

Enrichment results are passed to `exportFromSeurat()` via the
`enriched_pathways` parameter.

```{r, eval = FALSE}
exportFromSeurat(seurat_object,
  file = "my_data.crb",
  enriched_pathways = list(
    method = "cerebro_seurat_enrichr",
    results = enrichment_df
  )
)
```

See `vignette("overview_of_cerebro_v1.3_class")` for the expected format.

# See also

`vignette("overview_of_cerebro_v1.3_class")` for the Cerebro data model.
