---
title: "A real antigen-selected single-cell TCR demo, end to end"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{A real antigen-selected single-cell TCR demo, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# What this guide does

It records, step by step, how `demo_hla_tcr_dextramer.crb` is built from public 10x
Genomics data: what is downloaded, what each processing step changes, and what
the object finally declares. Every code chunk is the real pipeline, taken from
`data-raw/build_hla_tcr_dextramer_demo.R`; the chunks are not evaluated when the
vignette builds, because the raw inputs are ~1.6 GB.

The emphasis is on the two parts that decide whether the HLA & TCR Motifs page
says anything true: **how the TCR is transformed**, and **where the HLA comes
from**.

## Why this demo

One HLA demo ships, and everything in it is measured:

| | `demo_hla_tcr_dextramer.crb` |
|---|---|
| cells | **real** — 12,000 sequenced CD8+ T cells |
| TCR | **real** — paired αβ contigs, V gene + CDR3 |
| HLA | **real** — the donors' published genotypes (table S1), measured independently of these cells |

Earlier iterations shipped two others — a fully fabricated fixture and a real
bulk TCRβ cohort. Both are gone: cerebroAppLite is a single-cell application, and
a demo that is neither real nor single-cell earns its place only if nothing
better exists. This one is both, so it replaces them. Their build pipelines
remain in `data-raw/` (see `data-raw/hla.md`) if you want to reproduce them, and
the bulk workflow now lives in its own guide as *bring your own cohort*.

Replacing them required answering a fair objection first: *if the motif network
is only legible on synthetic data, what is the feature for?*

## The answer, as a measurement

A CDR3 Hamming-1 network needs an **antigen-selected** repertoire. An unselected
polyclonal repertoire is sparse in CDR3 space: neighbours at distance 1 are
rare, and adding cells does not fix it. A selected repertoire converges —
different donors independently arrive at near-identical CDR3s against the same
epitope (public / convergent recombination), which is exactly what the network
draws.

Measured on this source with the package's own motif core:

| subset of the same donor | unique CDR3β | result |
|---|---|---|
| all cells, unselected | 26,449 | trips the size guard — nothing to draw |
| dextramer-binding cells | 2,910 | **308 nodes in 75 motifs** |
| one epitope: Flu-MP `GILGFVFTL` | 267 | **121 nodes in 7 motifs** |

The last row is the point: against one immunodominant influenza epitope, 45% of
the observed CDR3s collapse into **seven** families. That is measured
convergence in real data, not a designed fixture.

# The source

10x Genomics, *CD8+ T cells of Healthy Donor 1–4* (2019) — the dextramer /
Immune Map experiment published as Zhang *et al.*, **Sci Adv** 7:eabf5835
(2021). CD8+ T cells from four HLA-haplotyped healthy donors were stained with a
pooled dCODE dextramer panel, sorted, and run on 10x 5′ immune profiling:
paired αβ TCR **and** transcriptome **and** surface protein **and** per-cell
dextramer counts. Licence: **CC BY 4.0**.

Three files per donor are used:

| file | ~size | what it carries |
|---|---|---|
| `..._all_contig_annotations.csv` | 34 MB | TCR contigs: barcode, chain, V/J gene, CDR3 |
| `..._binarized_matrix.csv` | 19–53 MB | per-cell dextramer calls, surface protein, donor |
| `..._filtered_feature_bc_matrix.tar.gz` | 296 MB | the gene-expression matrix |

# Download

```{r download}
# All four donors. The build script does this on demand and caches in
# data-raw/vdj_10x_dextramer/ (gitignored) -- use that exact directory, or the
# script will not find what you downloaded and will fetch it all again.
base <- "https://cf.10xgenomics.com/samples/cell-vdj/3.0.2"
cache <- "data-raw/vdj_10x_dextramer" # = CACHE in the build script
dir.create(cache, recursive = TRUE, showWarnings = FALSE)

for (d in 1:4) {
  stem <- sprintf("vdj_v1_hs_aggregated_donor%d", d)
  for (suffix in c(
    "_all_contig_annotations.csv",
    "_binarized_matrix.csv",
    "_filtered_feature_bc_matrix.tar.gz"
  )) {
    download.file(
      sprintf("%s/%s/%s%s", base, stem, stem, suffix),
      file.path(cache, paste0(stem, suffix)),
      mode = "wb"
    )
  }
}
```

Or in one line:

```bash
Rscript data-raw/build_hla_tcr_dextramer_demo.R
```

# The TCR, step by step

This is the half that decides what the network draws.

## What arrives

`all_contig_annotations.csv` is **one row per contig**, not per cell. A cell with
a paired receptor appears at least twice — once for TRA, once for TRB:

```
barcode              is_cell  chain  v_gene      j_gene     cdr3            productive
AAACCTGAGAAACCTA-1   True     TRA    TRAV12-2    TRAJ33     CAVNVAGKSTF     True
AAACCTGAGAAACCTA-1   True     TRB    TRBV19      TRBJ2-7    CASSIRSSYEQYF   True
```

## Step 1 — keep only productive αβ contigs

A non-productive contig carries a CDR3 that was never expressed; keeping it
would put a sequence in the network that the cell does not display.

```{r tcr-filter}
contigs <- read.csv(f$contigs, stringsAsFactors = FALSE)
contigs <- contigs[
  contigs$productive %in% c("True", "TRUE", TRUE) &
    contigs$chain %in% c("TRA", "TRB"),
  ,
  drop = FALSE
]
```

## Step 2 — contigs → one clonotype per cell

`scRepertoire::combineTCR()` collapses the contigs of a barcode into the single
row per cell that the app's parser expects. This is the **shape change** that
matters:

```{r tcr-combine}
out <- scRepertoire::combineTCR(
  list(contigs),
  samples = sprintf("donor%d", d),
  filterMulti = TRUE
)[[1]]
```

After it, one cell is one row and the chains live in underscore-joined strings:

```
barcode                      CTgene                                              CTaa
donor1_AAACCTGAGAAACCTA-1    TRAV12-2.TRAJ33.TRAC_TRBV19.TRBJ2-7.TRBC2           CAVNVAGKSTF_CASSIRSSYEQYF
```

`filterMulti = TRUE` keeps the dominant chain when a barcode reports more than
one — an ambiguous cell would otherwise contribute a CDR3 it may not carry.

## Why this exact shape

The app reads receptors through `hla_parse_ir_segments()`, whose contract is
`barcode` + `CTgene` + `CTaa`. For the requested chain it takes the matching
underscore slot, pulls the `TRBV…` / `TRBJ…` tokens out of `CTgene`, and the
CDR3 out of the same slot of `CTaa`.

**The V gene is required.** That is why the pipeline starts from the contigs and
not from the binarized matrix: the binarized matrix has the CDR3 amino acids but
no V/J gene at all, and this source identifies a receptor by *(V gene, CDR3)* —
declared as `receptor_key = "v_gene+cdr3"`, which makes split-by-V the app's
default so a node means what the source means.

# The HLA, step by step

This is the half that decides what the page may claim.

## What is genuinely real here

Each dextramer column is named `<allele>_<peptide>_<antigen>_binder`:

```
A0201_GILGFVFTL_Flu-MP_Influenza_binder
A0301_KLGGALQAK_IE-1_CMV_binder
A1101_AVFDRKSDAK_EBNA-3B_EBV_binder
```

The allele prefix is the **reagent's own HLA restriction** — a published
property of the dextramer, independent of these cells. So for every binding
event the pipeline recovers a real triple: *peptide, antigen, and the allele
that reagent is restricted by*.

```{r hla-parse}
allele_of <- function(col) {
  paste0("HLA-", sub("^([ABC])([0-9]{2})([0-9]{2})_.*", "\\1*\\2:\\3", col))
}
# A0201_GILGFVFTL_Flu-MP_Influenza_binder -> "HLA-A*02:01"
```

### What that triple is *not*

It is a property of the **reagent**, not a verified property of the cell. A
binder call does not establish that the cell is specific for that peptide, nor
that the reagent's allele is what presents it in that donor. Dextramer staining
is heavily cross-reactive in this experiment, and the size of the effect is
easy to state: for **6,729 of the 12,000 shipped cells (56 %)**, the bound
reagent's restriction is not even one of the alleles that donor carries
(per donor: 954, 468, 2,997 and 2,310 of 3,000) (see §*Why the
genotypes are not inferred from binding*, where the same fact sinks the
inference).

That is why the per-cell columns ship as `dextramer_antigen`,
`dextramer_peptide` and `dextramer_allele` — a reagent label, deliberately not
named `antigen` or `restricting_allele` — and why a fourth column,
`restriction_in_genotype` (`yes` / `no`), ships beside them. Colour the
projection by it and the noise is visible in one look rather than buried in a
methods note.

Nothing on the HLA Associations tab rests on these calls: the carrier contrasts
use the **published genotypes**, which were measured independently of these
cells.

## Step 3 — one binder call per cell, or none

10x's binarized calls mark each cell against each dextramer. Cells binding
several are **dropped, not guessed at**: an ambiguous call would place a cell in
the wrong HLA context, which is the one error this page must not make. Columns
containing `NR(` are the negative controls and are never evidence.

```{r hla-specificity}
hits <- as.matrix(b[, dex, drop = FALSE]) == "True"
hits[is.na(hits)] <- FALSE
keep <- rowSums(hits) == 1L                 # exactly one dextramer bound
idx <- max.col(hits, ties.method = "first")
```

Note what this filter does and does not buy: it removes *ambiguity*, not
*cross-reactivity*. A cell binding exactly one reagent is unambiguous, which is
not the same as being specific for it.

## Step 4 — donor genotype, from the published table

The donors' HLA haplotypes are published in **table S1** of the paper's
supplementary material. The link printed inside the paper is dead (the
`advances.sciencemag.org` domain was retired), but the file is served from
science.org:

```
https://www.science.org/doi/suppl/10.1126/sciadv.abf5835/suppl_file/abf5835_sm.pdf
```

Open it in a browser — science.org is behind Cloudflare and refuses
command-line clients. Nothing depends on that, though: the table is transcribed
inline in the build script, so it is self-contained — 14 rows, and a separate
file would only be one more thing to keep in step:

```{r hla-genotype}
DONOR_HLA <- read.csv(text = "donor,copy,allele
donor1,1,HLA-A*02:01
donor1,2,HLA-A*11:01
donor1,1,HLA-B*35:01
donor2,1,HLA-A*02:01
donor2,2,HLA-A*01:01
donor2,1,HLA-B*08:01
donor3,1,HLA-A*24:02
donor3,2,HLA-A*29:02
donor3,1,HLA-B*35:02
donor3,2,HLA-B*44:03
donor4,1,HLA-A*03:01
donor4,2,HLA-A*03:01
donor4,1,HLA-B*07:02
donor4,2,HLA-B*57:01", stringsAsFactors = FALSE)
```

| donor | HLA-A | HLA-B |
|---|---|---|
| Donor 1 | A\*02:01, A\*11:01 | B\*35:01 |
| Donor 2 | A\*02:01, A\*01:01 | B\*08:01 |
| Donor 3 | A\*24:02, A\*29:02 | B\*35:02, B\*44:03 |
| Donor 4 | A\*03:01, A\*03:01 | B\*07:02, B\*57:01 |

Because these were measured independently of the cells in this data set, a
carrier / non-carrier contrast on this demo is a real comparison.

## Why the genotypes are not inferred from binding

It is tempting to read the genotype off the data: a cell can only bind a
dextramer restricted by an allele its donor carries, so the binding profile
should reveal the haplotype. An earlier version of this build did exactly that,
requiring an allele to account for at least 200 cells **and** 10 % of a donor's
antigen-specific cells.

It was wrong. Against table S1:

| donor | inferred from binding | published |
|---|---|---|
| 3 | A\*03:01 | A\*24:02, A\*29:02 — **carries no A\*03:01** |

Donor 3 has **25,674 cells, 92.8 % of its antigen-specific cells**, binding
A\*03:01-restricted dextramers while carrying no A\*03:01 at all.
Cross-reactivity at that scale is not something a threshold separates, and the
inference would also have been circular: a donor would be a carrier *because*
their cells bound that allele's reagent, and the motifs come from those same
cells.

What the object still declares is the **selection**, not the genotype:

```{r hla-disclosure}
crb$technical_info$tcr_selection <- "antigen-selected"
# "... which receptors are present was decided by the panel. The donor HLA
#  genotypes are the published ones (table S1), measured independently of these
#  cells, so they are not circular with the selection."
crb$addHLATyping(donor_typing, source_type = "genotyped", ...)
```

# Cell selection — the step that makes the network legible

Only cells that (a) carry a clonotype with **both** chains and (b) bound
**exactly one** dextramer are kept. The dextramer sort is not a convenience; it
*is* the data set's defining property, and the reason the network draws at all
(see the table in §1).

The paired requirement is stricter than "has a `CTaa`", and deliberately so.
`combineTCR()` writes `CTaa` as `<alpha>_<beta>` and puts the literal string
`NA` on a side it could not resolve, so a non-empty `CTaa` may still be a single
chain. This demo is documented as paired αβ, so it has to actually be:

```{r selection}
sel <- merge(tcr_all, dex_all[dex_all$single_binder, ],
             by = c("barcode_raw", "donor"))

is_paired <- function(ctaa) {
  parts <- strsplit(ifelse(is.na(ctaa), "", ctaa), "_", fixed = TRUE)
  vapply(parts, function(p) {
    length(p) == 2L && all(nzchar(p)) && !any(p %in% c("NA", "None"))
  }, logical(1))
}
sel <- sel[is_paired(sel$CTaa), ]
# the deterministic per-donor subsample comes later, after the expression join,
# so the shipped object is exactly 3,000 cells per donor
```

# Expression and projection

Real measured transcriptomes, for exactly the cells that survived selection.
The matrix is subset **first**, so nothing is computed on cells that are thrown
away, and only the variable genes are shipped.

```{r gex}
m <- Seurat::Read10X(gex_dir)[["Gene Expression"]]
m <- m[, intersect(colnames(m), kept_barcodes), drop = FALSE]

so <- Seurat::CreateSeuratObject(counts = expr)
so <- Seurat::NormalizeData(so)
so <- Seurat::FindVariableFeatures(so, nfeatures = 2000)
so <- Seurat::ScaleData(so)
so <- Seurat::RunPCA(so, npcs = 30)
so <- Seurat::RunUMAP(so, dims = 1:30)

# Keep the shipped block SPARSE. Normalized single-cell expression is ~90%
# zeros; densifying this one cost 184 MiB of session memory and 4.5 MiB of
# installed package for nothing, and every other demo here is a dgCMatrix.
expression <- Seurat::GetAssayData(so, layer = "data")[hv, , drop = FALSE]
expression <- methods::as(expression, "CsparseMatrix")
```

# Assembling the `.crb`

```{r assemble}
crb <- Cerebro_v1.3$new()
crb$expression <- expression
crb$setMetaData(meta)          # cell_barcode, sample, cell_type,
                               # dextramer_antigen, dextramer_peptide,
                               # dextramer_allele, restriction_in_genotype
crb$projections <- list(umap = umap)
crb$immune_repertoire <- immune_repertoire   # split by donor
crb$technical_info <- list(
  observation_unit = "cell",
  receptor_key = "v_gene+cdr3",
  tcr_selection = "antigen-selected",
  tcr_selection_detail = "...",
  lineage_column = "cell_type"   # declared, so the app never has to guess
)
crb$addHLATyping(donor_typing, source_type = "genotyped", ...)
saveRDS(crb, "inst/extdata/v1.4/demo_hla_tcr_dextramer.crb", compress = "xz")
```

## The declared contracts, and why each one

| contract | value | why |
|---|---|---|
| `observation_unit` | `cell` | these really are sequenced cells, unlike the bulk demo |
| `receptor_key` | `v_gene+cdr3` | the source identifies a receptor by V **and** CDR3 |
| `tcr_selection` | `antigen-selected` | the repertoire was sorted for binding; the page must say so |
| `tcr_selection_detail` | see §6 | spells out the inferred-genotype circularity |
| `lineage_column` | `cell_type` | **declared**, so the app never infers it — every cell here is a sorted CD8 T cell |

Declaring `lineage_column` matters beyond tidiness: when it is absent the app
must guess which column holds the CD4/CD8 label, and a guess can land on a
treatment or study-arm column. Declaring it removes the guess.

# Verification: a gate, not a report

The build writes to a **staging** path first. It then re-reads that file and
re-derives the network with the package's own core, and every number it measures
is an **assertion**. Only if all of them hold does the staged file replace the
shipped `.crb`:

```{r verify}
saveRDS(crb, staged, compress = "xz")
check <- readRDS(staged)

stopifnot(
  "donors are not balanced"            = all(table(m$sample) == CELLS_PER_DONOR),
  "expression block is not sparse"     = inherits(check$expression, "CsparseMatrix"),
  "not every observation is paired"    = all(is_paired(ctaa)),
  "TRB network has collapsed"          = n_nodes > 100 && n_motifs >= 20,
  "HLA alleles drifted from table S1"  = setequal(paste(ht$sample, ht$allele), genotype_key)
  # ... plus provenance, projection alignment, and the honesty columns
)

file.rename(staged, OUT) # only reached if nothing above stopped the script
```

This is the difference between a build that *tells* you it broke and one that
*refuses to ship* broken: an earlier version printed these same numbers **after**
saving, so drifted inputs still replaced a good demo and still exited 0. The
thresholds are set well below the measured values — they detect drift, they are
not a transcript of one run.

One assertion is worth calling out because it looks backwards: the gate requires
that some cells still be **off-genotype**. If dextramer cross-reactivity ever
vanished from this data, the binder calls would have stopped being raw 10x calls
and every caveat in this vignette would need rewriting — so the caveat is
asserted rather than assumed.

On the shipped object the gate prints:

```
   cells: 12000 in 4 donors, balanced
   expression: dgCMatrix 2000x12000, 40.2 MiB in memory
   repertoire: 12000 observations, all paired; chains TRA, TRB
   TRB: 3270 unique CDR3 -> 169 nodes in 39 motifs
   TRA: 3189 unique CDR3 -> 396 nodes in 141 motifs
   HLA: 4 donors, 12 alleles, source_type=genotyped
   off-genotype binder calls: 6729 of 12000 cells (the documented caveat)
   PUBLISHED inst/extdata/v1.4/demo_hla_tcr_dextramer.crb (5.2 MB)
```

12,000 dextramer-selected cells (3,000 per donor, every one paired αβ), 2,000
variable genes, **5.2 MB** — it was 7.9 MB while the expression block was
stored dense.

# What you see in the app

Open the data set and go to **HLA & TCR Motifs**:

* **Motif Network** — families of real CDR3β sequences at Hamming distance 1.
  Colour by *Sample of origin* to see the same motif reached independently by
  more than one donor: that is convergence, and it is the whole argument.
* **Network data** — the rows behind the picture. Besides the structural columns
  it carries whatever the data set declares as a grouping, so here that includes
  each node's `dextramer_antigen`, `dextramer_allele` and
  `restriction_in_genotype` — the reagent it bound, and whether that reagent's
  restriction is an allele the donor actually carries.
* **HLA Associations** — real donor genotypes, so the carrier contrasts are
  real; read the note above the tables for what the antigen selection means.
  These contrasts do **not** use the dextramer binder calls.
* **Data & QC** — the published genotypes and their coverage.
