---
title: "A real antigen-selected single-cell TCR demo, end to end"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{A real antigen-selected single-cell TCR demo, end to end}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

# What this guide does

It records, step by step, how `demo_hla_tcr_dextramer.crb` is built from public 10x
Genomics data: what is downloaded, what each processing step changes, and what
the object finally declares. Every code chunk is the real pipeline, taken from
`data-raw/build_hla_tcr_dextramer_demo.R`; the chunks are not evaluated when the
vignette builds, because the raw inputs are ~1.6 GB.

The emphasis is on the two parts that decide whether the HLA & TCR Motifs page
says anything true: **how the TCR is transformed**, and **where the HLA comes
from**.

## Why a third demo

The package ships three HLA demos because no single one answers everything:

| demo | cells | TCR | HLA | what it is for |
|---|---|---|---|---|
| `demo_hla_tcr_synthetic.crb` | synthetic | synthetic | synthetic | shows a dense motif network; proves nothing about real data |
| `demo_hla_tcr_bulk.crb` | none (bulk) | **real** | **real genotypes** | HLA Associations on genuine typing |
| **`demo_hla_tcr_dextramer.crb`** | **real** | **real** | inferred (see §6) | the motif network on measured sequences |

The gap the third one fills is a fair objection: *if the motif network is only
legible on synthetic data, what is the feature for?*

## The answer, as a measurement

A CDR3 Hamming-1 network needs an **antigen-selected** repertoire. An unselected
polyclonal repertoire is sparse in CDR3 space: neighbours at distance 1 are
rare, and adding cells does not fix it. A selected repertoire converges —
different donors independently arrive at near-identical CDR3s against the same
epitope (public / convergent recombination), which is exactly what the network
draws.

Measured on this source with the package's own motif core:

| subset of the same donor | unique CDR3β | result |
|---|---|---|
| all cells, unselected | 26,449 | trips the size guard — nothing to draw |
| dextramer-binding cells | 2,910 | **308 nodes in 75 motifs** |
| one epitope: Flu-MP `GILGFVFTL` | 267 | **121 nodes in 7 motifs** |

The last row is the point: against one immunodominant influenza epitope, 45% of
the observed CDR3s collapse into **seven** families. That is measured
convergence in real data, not a designed fixture.

# The source

10x Genomics, *CD8+ T cells of Healthy Donor 1–4* (2019) — the dextramer /
Immune Map experiment published as Zhang *et al.*, **Sci Adv** 7:eabf5835
(2021). CD8+ T cells from four HLA-haplotyped healthy donors were stained with a
pooled dCODE dextramer panel, sorted, and run on 10x 5′ immune profiling:
paired αβ TCR **and** transcriptome **and** surface protein **and** per-cell
dextramer counts. Licence: **CC BY 4.0**.

Three files per donor are used:

| file | ~size | what it carries |
|---|---|---|
| `..._all_contig_annotations.csv` | 34 MB | TCR contigs: barcode, chain, V/J gene, CDR3 |
| `..._binarized_matrix.csv` | 19–53 MB | per-cell dextramer calls, surface protein, donor |
| `..._filtered_feature_bc_matrix.tar.gz` | 296 MB | the gene-expression matrix |

# Download

```{r download}
# All four donors; the build script does this on demand and caches in
# data-raw/vdj_10x_dextramer/ (gitignored). Manually:
base <- "https://cf.10xgenomics.com/samples/cell-vdj/3.0.2"
dir.create("data-raw/vdj_10x", recursive = TRUE, showWarnings = FALSE)

for (d in 1:4) {
  stem <- sprintf("vdj_v1_hs_aggregated_donor%d", d)
  for (suffix in c(
    "_all_contig_annotations.csv",
    "_binarized_matrix.csv",
    "_filtered_feature_bc_matrix.tar.gz"
  )) {
    download.file(
      sprintf("%s/%s/%s%s", base, stem, stem, suffix),
      file.path("data-raw/vdj_10x", paste0(stem, suffix)),
      mode = "wb"
    )
  }
}
```

Or in one line:

```bash
Rscript data-raw/build_hla_tcr_dextramer_demo.R
```

# The TCR, step by step

This is the half that decides what the network draws.

## What arrives

`all_contig_annotations.csv` is **one row per contig**, not per cell. A cell with
a paired receptor appears at least twice — once for TRA, once for TRB:

```
barcode              is_cell  chain  v_gene      j_gene     cdr3            productive
AAACCTGAGAAACCTA-1   True     TRA    TRAV12-2    TRAJ33     CAVNVAGKSTF     True
AAACCTGAGAAACCTA-1   True     TRB    TRBV19      TRBJ2-7    CASSIRSSYEQYF   True
```

## Step 1 — keep only productive αβ contigs

A non-productive contig carries a CDR3 that was never expressed; keeping it
would put a sequence in the network that the cell does not display.

```{r tcr-filter}
contigs <- read.csv(f$contigs, stringsAsFactors = FALSE)
contigs <- contigs[
  contigs$productive %in% c("True", "TRUE", TRUE) &
    contigs$chain %in% c("TRA", "TRB"),
  ,
  drop = FALSE
]
```

## Step 2 — contigs → one clonotype per cell

`scRepertoire::combineTCR()` collapses the contigs of a barcode into the single
row per cell that the app's parser expects. This is the **shape change** that
matters:

```{r tcr-combine}
out <- scRepertoire::combineTCR(
  list(contigs),
  samples = sprintf("donor%d", d),
  filterMulti = TRUE
)[[1]]
```

After it, one cell is one row and the chains live in underscore-joined strings:

```
barcode                      CTgene                                              CTaa
donor1_AAACCTGAGAAACCTA-1    TRAV12-2.TRAJ33.TRAC_TRBV19.TRBJ2-7.TRBC2           CAVNVAGKSTF_CASSIRSSYEQYF
```

`filterMulti = TRUE` keeps the dominant chain when a barcode reports more than
one — an ambiguous cell would otherwise contribute a CDR3 it may not carry.

## Why this exact shape

The app reads receptors through `hla_parse_ir_segments()`, whose contract is
`barcode` + `CTgene` + `CTaa`. For the requested chain it takes the matching
underscore slot, pulls the `TRBV…` / `TRBJ…` tokens out of `CTgene`, and the
CDR3 out of the same slot of `CTaa`.

**The V gene is required.** That is why the pipeline starts from the contigs and
not from the binarized matrix: the binarized matrix has the CDR3 amino acids but
no V/J gene at all, and this source identifies a receptor by *(V gene, CDR3)* —
declared as `receptor_key = "v_gene+cdr3"`, which makes split-by-V the app's
default so a node means what the source means.

# The HLA, step by step

This is the half that decides what the page may claim.

## What is genuinely real here

Each dextramer column is named `<allele>_<peptide>_<antigen>_binder`:

```
A0201_GILGFVFTL_Flu-MP_Influenza_binder
A0301_KLGGALQAK_IE-1_CMV_binder
A1101_AVFDRKSDAK_EBNA-3B_EBV_binder
```

The allele prefix is the **reagent's own HLA restriction** — a published
property of the dextramer, independent of these cells. So for every
antigen-specific cell the pipeline recovers a real triple: *peptide, antigen,
and the HLA allele that presents it*.

```{r hla-parse}
allele_of <- function(col) {
  paste0("HLA-", sub("^([ABC])([0-9]{2})([0-9]{2})_.*", "\\1*\\2:\\3", col))
}
# A0201_GILGFVFTL_Flu-MP_Influenza_binder -> "HLA-A*02:01"
```

## Step 3 — one specificity per cell, or none

10x's binarized calls mark each cell against each dextramer. Cells binding
several are **dropped, not guessed at**: an ambiguous specificity would place a
cell in the wrong HLA context, which is the one error this page must not make.
Columns containing `NR(` are the negative controls and are never evidence.

```{r hla-specificity}
hits <- as.matrix(b[, dex, drop = FALSE]) == "True"
hits[is.na(hits)] <- FALSE
keep <- rowSums(hits) == 1L                 # exactly one specificity
idx <- max.col(hits, ties.method = "first")
```

## Step 4 — donor genotype, and the honest caveat

The published haplotypes are in table S1, served from
`advances.sciencemag.org` — a domain retired when Science migrated to
science.org, so the supplement is no longer retrievable. The build therefore
**infers** each donor's alleles from which dextramers their cells bound.

A count threshold alone is not enough. Donor 2 has 814 cells against
HLA-A\*11:01, which clears any sensible count, and that is still only **2 %** of
that donor's antigen-specific cells — cross-reactivity, not a genotype. The cut
is therefore relative:

```{r hla-genotype}
CARRIER_MIN_CELLS <- 200L    # absolute floor
CARRIER_MIN_SHARE <- 0.10    # ... and a real share of the donor's specific cells
counts <- tab[donor, ]
share <- counts / sum(counts)
alleles <- names(counts)[counts >= CARRIER_MIN_CELLS & share >= CARRIER_MIN_SHARE]
```

That reproduces the per-donor allele profile of the paper's own
quality-controlled call set (data file S1) exactly, and it leaves the four
donors **not** identical — which is what makes a carrier / non-carrier split
exist at all:

| allele | donor1 | donor2 | donor3 | donor4 | contrast |
|---|---|---|---|---|---|
| HLA-A*02:01 | ✓ | ✓ | · | · | 2 vs 2 |
| HLA-A*03:01 | ✓ | ✓ | ✓ | ✓ | none |
| HLA-A*11:01 | ✓ | · | · | ✓ | 2 vs 2 |
| HLA-B*08:01 | · | ✓ | · | · | 1 vs 3 |

## ⚠️ The circularity, stated plainly

A donor is called a carrier of HLA-X **because their cells bound an
X-restricted dextramer** — and the motif families are built from those same
cells. So any carrier / non-carrier contrast this data set shows is **guaranteed
by construction and is not independent evidence** of an HLA association.

The object says so itself, through the contract the app reads:

```{r hla-disclosure}
crb$technical_info$tcr_selection <- "antigen-selected"
crb$technical_info$tcr_selection_detail <- paste(
  "... the donor genotypes shipped here are INFERRED from which dextramers",
  "each donor's cells bound ... any carrier / non-carrier contrast on this",
  "data set is circular and is NOT independent evidence."
)
crb$addHLATyping(
  donor_typing,
  source_type = "imputed",
  typing_method = "inferred from dCODE dextramer binding (not a genotyping assay)"
)
```

The app prints that detail above the Associations tables. For association work
on genuine, independently measured genotypes, use `demo_hla_tcr_bulk.crb`.

If table S1 ever becomes available again, swapping in the published table and
setting `source_type = "genotyped"` is the only change needed — and the
circularity note above can then be dropped.

# Cell selection — the step that makes the network legible

Only cells that (a) carry a productive clonotype and (b) bound **exactly one**
dextramer are kept. This is not a convenience; it *is* the data set's defining
property, and the reason the network draws at all (see the table in §1).

```{r selection}
sel <- merge(tcr_all, dex_all[dex_all$specific, ],
             by = c("barcode_raw", "donor"))
sel <- sel[!is.na(sel$CTaa) & nzchar(sel$CTaa), ]
# then a deterministic per-donor subsample, so the shipped file stays small
```

# Expression and projection

Real measured transcriptomes, for exactly the cells that survived selection.
The matrix is subset **first**, so nothing is computed on cells that are thrown
away, and only the variable genes are shipped.

```{r gex}
m <- Seurat::Read10X(gex_dir)[["Gene Expression"]]
m <- m[, intersect(colnames(m), kept_barcodes), drop = FALSE]

so <- Seurat::CreateSeuratObject(counts = expr)
so <- Seurat::NormalizeData(so)
so <- Seurat::FindVariableFeatures(so, nfeatures = 2000)
so <- Seurat::ScaleData(so)
so <- Seurat::RunPCA(so, npcs = 30)
so <- Seurat::RunUMAP(so, dims = 1:30)
```

# Assembling the `.crb`

```{r assemble}
crb <- Cerebro_v1.3$new()
crb$expression <- expression
crb$setMetaData(meta)          # cell_barcode, sample, cell_type, antigen,
                               # peptide, restricting_allele
crb$projections <- list(umap = umap)
crb$immune_repertoire <- immune_repertoire   # split by donor
crb$technical_info <- list(
  observation_unit = "cell",
  receptor_key = "v_gene+cdr3",
  tcr_selection = "antigen-selected",
  tcr_selection_detail = "...",
  lineage_column = "cell_type"   # declared, so the app never has to guess
)
crb$addHLATyping(donor_typing, source_type = "imputed", ...)
saveRDS(crb, "inst/extdata/v1.4/demo_hla_tcr_dextramer.crb", compress = "xz")
```

## The declared contracts, and why each one

| contract | value | why |
|---|---|---|
| `observation_unit` | `cell` | these really are sequenced cells, unlike the bulk demo |
| `receptor_key` | `v_gene+cdr3` | the source identifies a receptor by V **and** CDR3 |
| `tcr_selection` | `antigen-selected` | the repertoire was sorted for binding; the page must say so |
| `tcr_selection_detail` | see §6 | spells out the inferred-genotype circularity |
| `lineage_column` | `cell_type` | **declared**, so the app never infers it — every cell here is a sorted CD8 T cell |

Declaring `lineage_column` matters beyond tidiness: when it is absent the app
must guess which column holds the CD4/CD8 label, and a guess can land on a
treatment or study-arm column. Declaring it removes the guess.

# Verification: measured, never assumed

The build script ends by re-reading the file it just wrote and re-deriving the
network with the package's own core, so the numbers reported are what the
shipped object actually produces:

```{r verify}
check <- readRDS(OUT)
ir <- check$getImmuneRepertoire()
for (ch in c("TRB", "TRA")) {
  seg <- cerebroAppLite:::hla_parse_ir_segments(ir, ch)
  g <- cerebroAppLite:::hla_build_motif_graph(seg, by_v = TRUE, min_nodes = 2L)
  # -> unique CDR3, nodes, motifs
}
```

A build that silently loses the motif structure therefore reports it, instead of
shipping a data set whose network is empty. On the shipped object it prints:

```
   chains: TRA, TRB
   TRB: 3350 unique CDR3 -> 157 nodes in 31 motifs
   TRA: 3067 unique CDR3 -> 367 nodes in 130 motifs
   HLA: 4 donors, 4 alleles, source_type=imputed
```

12,000 antigen-selected cells (3,000 per donor), 2,000 variable genes, 7.8 MB.

# What you see in the app

Open the data set and go to **HLA & TCR Motifs**:

* **Motif Network** — families of real CDR3β sequences at Hamming distance 1.
  Colour by *Sample of origin* to see the same motif reached independently by
  more than one donor: that is convergence, and it is the whole argument.
* **Network data** — the rows behind the picture, including each node's antigen
  and the HLA allele that presents it.
* **HLA Associations** — works, but read the disclosure above the tables first:
  on this data set the contrast is circular by construction.
* **Data & QC** — the inferred genotypes and their coverage.
