---
title: "Immune Repertoire Analysis in cerebroAppLite"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Immune Repertoire Analysis in cerebroAppLite}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Overview

cerebroAppLite v1.4 introduces a comprehensive **Immune Repertoire** module for
interactive exploration of T-cell and B-cell receptor clonotype data. The module
leverages the `scRepertoire` package and supports 19 visualization methods covering
clonal abundance, diversity, CDR3 properties, gene usage, and cross-sample
comparisons.

The Immune Repertoire tab appears **conditionally** in the sidebar — only when
the loaded `.crb` file contains TCR or BCR clonotype annotations. The bundled
`example.crb` ships with real 10x immune-repertoire data (the
`sc5p_v2_hs_PBMC_10k` dataset, with 5' gene expression, TCR, and BCR from the
same experiment), so the tab — including TCR, BCR (isotype/SHM), and
cross-sample comparisons — is available out of the box. (This single 10x donor
is randomly split into three demo samples so cross-sample features have data;
the sample labels are not distinct biological donors.)

Grouping and sample-splitting in the module work for **any** grouping variable
present in the data set's cell metadata (sample, condition, cell type, etc.).
The clonotype tables themselves only need the standard scRepertoire columns;
the module joins metadata onto them by barcode at runtime.

# Preparing immune repertoire data

## Background: scRepertoire

The immune repertoire module is built on the
[scRepertoire](https://www.borch.dev/uploads/screpertoire/) package (≥ 2.0),
which is the standard tool for turning 10x Cell Ranger V(D)J output into
per-cell clonotype annotations. cerebroAppLite lists it under `Suggests`, so
install it once before preparing data:

```{r, eval = FALSE}
# Bioconductor
BiocManager::install("scRepertoire")
```

The full scRepertoire workflow is documented in its vignettes —
[Loading data](https://www.borch.dev/uploads/screpertoire/articles/loading) and
[Combining contigs](https://www.borch.dev/uploads/screpertoire/articles/combining_contigs).
The steps below summarise just what cerebroAppLite needs.

## Step 1 — load the V(D)J contigs

Read the `filtered_contig_annotations.csv` produced by Cell Ranger for each
sample. TCR and BCR are separate libraries, so load each one:

```{r, eval = FALSE}
library(scRepertoire)

# one filtered_contig_annotations.csv per sample
tcr_contigs <- lapply(tcr_paths, read.csv) # T-cell receptor contigs
bcr_contigs <- lapply(bcr_paths, read.csv) # B-cell receptor contigs
```

## Step 2 — combine contigs into clonotypes

`combineTCR()` and `combineBCR()` collapse contigs into per-cell clonotypes and
add the standard `CTgene` / `CTnt` / `CTaa` / `CTstrict` columns. Use the
matching function for each receptor type:

```{r, eval = FALSE}
combined_tcr <- combineTCR(tcr_contigs, samples = sample_names)
combined_bcr <- combineBCR(bcr_contigs, samples = sample_names)
```

## Step 3 — attach to the Seurat object and export

Build a named list of data.frames (one element per sample) holding the
scRepertoire clonotype columns, attach it to the Seurat object's
`@misc$immune_repertoire` slot, and export. `exportFromSeurat()` picks the slot
up automatically. To include both receptor types, row-bind the T- and B-cell
clonotypes per sample (cells are mutually exclusive — a cell has a TCR *or* a
BCR):

```{r, eval = FALSE}
library(cerebroAppLite)

ir_cols <- c("barcode", "CTgene", "CTnt", "CTaa", "CTstrict")
to_df <- function(x) x[, ir_cols, drop = FALSE]

ir_data <- lapply(sample_names, function(s) {
  rbind(to_df(combined_tcr[[s]]), to_df(combined_bcr[[s]]))
})
names(ir_data) <- sample_names

seurat_object@misc$immune_repertoire <- ir_data

exportFromSeurat(
  seurat_object,
  file = "my_data.crb",
  experiment_name = "my_experiment",
  organism = "hg",
  groups = c("sample", "seurat_clusters")
)
```

Each data.frame must contain the scRepertoire clonotype columns `barcode`,
`CTgene`, `CTnt`, `CTaa`, and `CTstrict`. The `barcode` values should match
the cell barcodes in the Seurat object's metadata. Chains (e.g. TRA/TRB,
IGH/IGK/IGL) are detected automatically from `CTgene`, and grouping variables
(sample, condition, cell type, ...) are read from the cell metadata at runtime —
the clonotype tables themselves only need the five columns above.

# Module interface

## Settings panel

The settings panel provides three controls:

- **Clone call**: which clonotype identifier to use for analysis
  (`gene`, `nt`, `aa`, or `strict`)
- **Group by**: a metadata grouping variable for faceted comparisons
- **Chain**: filter to TCR chains (TRA, TRB, TRG, TRD), BCR chains
  (IGH, IGK, IGL), or all (`both`)

When multiple samples are present, additional controls appear for scatter
plot sample selection and multi-sample comparison.

## Visualization tabs

The module provides 19 tabbed visualizations. Each includes a contextual help
panel explaining the biological interpretation with example guidance.

### Basic repertoire metrics

- **Abundance**: ranks clonotypes by cell count; steep drop-off indicates
  oligoclonal dominance
- **Clonal UMAP**: overlays clone-expansion level
  (Single/Small/Medium/Large/Hyperexpanded) on the cell projection, reusing the
  dataset's UMAP/tSNE coordinates; a Receptor (TCR/BCR) and Projection selector
  drive it
- **Diversity**: Shannon entropy with bootstrap confidence intervals
- **Homeostasis**: categorizes clonotypes into size classes from Rare to
  Hyperexpanded
- **Proportion**: cumulative fraction of the repertoire occupied by top-ranked
  clonotypes

### Sequence properties

- **Length**: CDR3 length distribution; shifts suggest antigen-driven selection
- **AA %**: positional amino acid composition at each CDR3 position
- **Entropy**: Shannon entropy at each CDR3 position
- **Property**: physicochemical profiles (Atchley, Kidera, and other scales)
- **K-mer**: top recurring short amino acid motifs

### Gene usage

- **Gene usage / vizGenes / percentGenes**: V(D)J gene segment usage
  frequencies
- **percentVJ**: V-J gene pairing frequencies

### Cross-sample analysis

When >= 2 samples are present, additional tabs become available:
- **Compare**: alluvial diagram tracking clonotypes across samples
- **Overlap**: pairwise clonotype sharing heatmap
- **Scatter**: clone frequency comparison between two samples
- **SizeDist**: hierarchical clustering of samples by clone size distribution

### Quality control

- **Quant**: total unique clonotype count per sample
- **Rarefaction**: clonotype discovery curves for sequencing saturation assessment
