---
title: "Most Expressed Genes"
author: "Xuesong Wang"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Most Expressed Genes}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

# Overview

The **Most expressed genes** tab shows, for every group in the dataset, the
genes with the highest mean expression. Unlike marker genes (which identify
genes that differ _between_ groups), most expressed genes simply rank genes
by their average expression _within_ each group.

This tab is always visible in the sidebar because expression data is present
in every `.crb` file.

# Quick start

```{r, eval = FALSE}
library(CerebroNexus)
launchCerebroV1.4()
```

1. Launch CerebroNexus and load a `.crb` file
2. Click **Most expressed genes** in the sidebar
3. Select a grouping variable (e.g., `seurat_clusters`, `sample`)
4. The table shows, for each group, the gene name, mean expression, and
   percentage of cells expressing that gene

# Table features

The results table is rendered as an interactive DT table:

- **Search**: filter genes by name across all groups
- **Sort**: click column headers to sort by expression, percentage, or group
- **Download**: use the export buttons to save results as CSV or Excel

# Data export

Most expressed gene tables are computed automatically during
`exportFromSeurat()`. No additional parameters are required.

```{r, eval = FALSE}
exportFromSeurat(seurat_object, file = "my_data.crb")
```

# See also

`vignette("cerebroApp_workflow_Seurat")` for the complete export workflow.
