---
title: "Testing procedure"
author: "Roman Hillje"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    number_sections: true
    self_contained: yes
    mode: selfcontained
vignette: >
  %\VignetteIndexEntry{Testing procedure}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

# Preparation

## Load packages

```{r eval=FALSE}
library(dplyr)
library(shiny)
library(shinydashboard)
```

```{r eval=FALSE}
library(Seurat)
library(cerebroAppLite)
```

## Classes

### `Group`

```{r eval=FALSE}
source('R/class-Group.R')
```

```{r eval=FALSE}
Group$new(
  name = 'sample',
  levels = levels(cerebro_seurat$meta_data$sample),
  description = 'experimental samples',
  colors = c('black','red','green')
)
```

```{r eval=FALSE}
Group$new(
  name = 'sample',
  levels = levels(cerebro_seurat$meta_data$sample),
  description = 'experimental samples',
  colors = c('black','red')
)
```

### `GeneList`

```{r eval=FALSE}
source('R/class-GeneList.R')
```

```{r eval=FALSE}
GeneList$new(
  name = 'test',
  genes = c('a','b','c'),
  description = 'genes to assign cell cycle'
)
```

```{r eval=FALSE}
GeneList$new(
  name = 'test',
  genes = 1,
  description = 'genes to assign cell cycle'
)
```

```{r eval=FALSE}
GeneList$new(
  name = 'test',
  genes = c('a','b')
)
```

### `MetaData`

```{r eval=FALSE}
source('R/class-MetaData.R')
```

```{r eval=FALSE}
table <- Table$new(
  name = 'test',
  table = cerebro_seurat$marker_genes$cerebro_seurat$sample,
  description = 'this is a description'
)
table
```

### `Table`

```{r eval=FALSE}
source('R/class-Table.R')
```

```{r eval=FALSE}
table <- Table$new(
  name = 'test',
  table = cerebro_seurat$marker_genes$cerebro_seurat$sample,
  description = 'this is a description'
)
table
```

### `Projection`

```{r eval=FALSE}
source('R/class-Projection.R')
```

```{r eval=FALSE}
projection <- Projection$new(
  name = 'test',
  dimensions = 2,
  coordinates = cerebro_seurat$projections$UMAP,
  description = 'this is a description'
)
projection
```

### `Tree`

```{r eval=FALSE}
source('R/class-Tree.R')
```

```{r eval=FALSE}
tree <- Tree$new(
  group = 'sample',
  tree = cerebro_seurat$trees$sample,
  description = 'this tree shows the relationship between samples'
)
tree
```

## Load data

```{r eval=FALSE}
pbmc_Seurat <- readRDS('~/Dropbox/GSE_stuff/data/seurat.rds')
```

# Process data

## Dimensional reduction

```{r eval=FALSE}
pbmc_Seurat <- RunUMAP(
  pbmc_Seurat,
  reduction.name = 'UMAP_3D',
  reduction.key = 'UMAP3D_',
  dims = 1:30,
  n.components = 3,
  seed.use = 100,
  verbose = FALSE
)
```

## Trees

```{r eval=FALSE}
Idents(pbmc_Seurat) <- "sample"
pbmc_Seurat <- BuildClusterTree(
  pbmc_Seurat,
  dims = 1:30,
  reorder = FALSE,
  reorder.numeric = FALSE,
  verbose = FALSE
)
pbmc_Seurat@misc$trees$sample <- pbmc_Seurat@tools$BuildClusterTree

Idents(pbmc_Seurat) <- "seurat_clusters"
pbmc_Seurat <- BuildClusterTree(
  pbmc_Seurat,
  dims = 1:30,
  reorder = FALSE,
  reorder.numeric = FALSE,
  verbose = FALSE
)
pbmc_Seurat@misc$trees$seurat_clusters <- pbmc_Seurat@tools$BuildClusterTree

Idents(pbmc_Seurat) <- "cell_type_singler_blueprintencode_main"
pbmc_Seurat <- BuildClusterTree(
  pbmc_Seurat,
  dims = 1:30,
  reorder = FALSE,
  reorder.numeric = FALSE,
  verbose = FALSE
)
pbmc_Seurat@misc$trees$cell_type_singler_blueprintencode_main <- pbmc_Seurat@tools$BuildClusterTree
```

## Marker genes

```{r eval=FALSE}
source('~/Research/GitHub/cerebroAppLite/R/getMarkerGenes.R')
```

```{r eval=FALSE}
pbmc_Seurat@misc$marker_genes$by_sample <- NULL
pbmc_Seurat@misc$marker_genes$by_cluster <- NULL
pbmc_Seurat@misc$marker_genes$parameters <- NULL

pbmc_Seurat <- getMarkerGenes(
  pbmc_Seurat,
  assay = 'RNA',
  organism = 'hg',
  groups = c('sample','seurat_clusters','cell_type_singler_blueprintencode_main'),
  name = 'cerebro_seurat',
  only_pos = TRUE,
  min_pct = 0.7,
  thresh_logFC = 0.25,
  thresh_p_val = 0.01,
  test = 'wilcox',
  verbose = TRUE
)
```

## Meta data

```{r eval=FALSE}
pbmc_Seurat@misc$technical_info$cerebroApp_version <- utils::packageVersion('cerebroAppLite')
pbmc_Seurat@misc$technical_info$Seurat <- utils::packageVersion('Seurat')
```

# Export data

```{r eval=FALSE}
source('~/Research/GitHub/cerebroAppLite/R/class-Cerebro.R')

source('~/Research/GitHub/cerebroAppLite/R/exportFromSeurat.R')
source('~/Research/GitHub/cerebroAppLite/R/exportFromSCE.R')
```

## `dgCMatrix`

```{r eval=FALSE}
exportFromSeurat(
  pbmc_Seurat,
  assay = 'SCT',
  slot = 'data',
  file = '~/Dropbox/Cerebro_development/pbmc_Seurat_dgCMatrix.crb',
  experiment_name = 'pbmc_Seurat',
  organism = 'hg',
  groups = c('sample','seurat_clusters','cell_type_singler_blueprintencode_main'),
  cell_cycle = c('cell_cycle_seurat'),
  nUMI = 'nCount_RNA',
  nGene = 'nFeature_RNA',
  add_all_meta_data = TRUE,
  use_delayed_array = FALSE,
  verbose = TRUE
)
```

## `RleMatrix`

```{r eval=FALSE}
exportFromSeurat(
  pbmc_Seurat,
  assay = 'SCT',
  slot = 'data',
  file = '~/Dropbox/Cerebro_development/pbmc_Seurat_RleMatrix.crb',
  experiment_name = 'pbmc_Seurat',
  organism = 'hg',
  groups = c('sample','seurat_clusters','cell_type_singler_blueprintencode_main'),
  cell_cycle = c('cell_cycle_seurat'),
  nUMI = 'nCount_RNA',
  nGene = 'nFeature_RNA',
  add_all_meta_data = TRUE,
  use_delayed_array = TRUE,
  verbose = TRUE
)
```

## BPCells (Seurat v5)

```{r eval=FALSE}
exportFromSeurat(
  pbmc_Seurat,
  assay = 'RNA',
  slot = 'data',
  file = '~/Dropbox/Cerebro_development/pbmc_Seurat_BPCells.crb',
  experiment_name = 'pbmc_Seurat',
  organism = 'hg',
  groups = c('sample','seurat_clusters','cell_type_singler_blueprintencode_main'),
  cell_cycle = c('cell_cycle_seurat'),
  nUMI = 'nCount_RNA',
  nGene = 'nFeature_RNA',
  add_all_meta_data = TRUE,
  use_delayed_array = FALSE,
  expression_matrix_BPCells = TRUE,
  verbose = TRUE
)
```

# Load data to check

```{r eval=FALSE}
cerebro_seurat <- readRDS('~/Dropbox/Cerebro_development/pbmc_Seurat_dgCMatrix.crb')
cerebro_seurat
# class: Cerebro
# cerebroAppLite version: 1.5.0
# experiment name: pbmc_Seurat
# organism: hg
# date of analysis: 2020-02-19
# date of export: 2020-08-20
# number of cells: 5,697
# number of genes: 15,907
# grouping variables (3): sample, seurat_clusters, cell_type_singler_blueprintencode_main
# cell cycle variables (1): cell_cycle_seurat
# projections (2): UMAP, UMAP_3D
# trees (3): sample, seurat_clusters, cell_type_singler_blueprintencode_main
# marker genes:
#   - cerebro_seurat (3): sample, seurat_clusters, cell_type_singler_blueprintencode_main, 
#   - test (1): sample
```

# Launch Cerebro

## Set options

### Pre-load with `dgCMatrix`

```{r eval=FALSE}
Cerebro.options <- list(
  "mode" = "open",
  "crb_file_to_load" = "~/Dropbox/Cerebro_development/pbmc_Seurat_dgCMatrix.crb",
  "cerebro_root" = "~/Research/GitHub/cerebroAppLite/inst/"
)
```

### Pre-load with `RleMatrix`

```{r eval=FALSE}
Cerebro.options <- list(
  "mode" = "open",
  "crb_file_to_load" = "~/Dropbox/Cerebro_development/pbmc_Seurat_RleMatrix.crb",
  "cerebro_root" = "~/Research/GitHub/cerebroAppLite/inst/"
)
```

### Pre-load with BPCells (Seurat v5)

```{r eval=FALSE}
Cerebro.options <- list(
  "mode" = "open",
  "crb_file_to_load" = "~/Dropbox/Cerebro_development/pbmc_Seurat_BPCells.crb",
  "cerebro_root" = "~/Research/GitHub/cerebroAppLite/inst/"
)
```

### Pre-load with example data set

```{r eval=FALSE}
Cerebro.options <- list(
  "mode" = "open",
  "crb_file_to_load" = "~/Research/GitHub/cerebroAppLite/inst/extdata/example.crb",
  "cerebro_root" = "~/Research/GitHub/cerebroAppLite/inst"
)
```

### No pre-loading

```{r eval=FALSE}
Cerebro.options <- list(
  "mode" = "open",
  "cerebro_root" = "~/Research/GitHub/cerebroAppLite/inst"
)
```

## Start Cerebro

### Without profvis

```{r eval=FALSE}
options(shiny.maxRequestSize = 800 * 1024^2)

source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_UI.R")
source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_server.R")

shiny::shinyApp(ui = ui, server = server)
```

### With provfis

```{r eval=FALSE}
options(shiny.maxRequestSize = 800 * 1024^2)

source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_UI.R")
source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_server.R")

profvis::profvis(shiny::runApp(shiny::shinyApp(ui = ui, server = server)))
```

### With reactlog

```{r eval=FALSE}
library(reactlog)

reactlog_enable()

options(shiny.maxRequestSize = 800 * 1024^2)

source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_UI.R")
source("~/Research/GitHub/cerebroAppLite/inst/shiny/shiny_server.R")

shiny::shinyApp(ui = ui, server = server)
```

```{r eval=FALSE}
shiny::reactlogShow()
```

# Install `cerebroAppLite`

```{r eval=FALSE}
devtools::document()
devtools::document()

devtools::install_local(
  '.',
  force = TRUE,
  upgrade = FALSE,
  build_vignettes = FALSE
)

cerebroAppLite::launchCerebro(
  crb_file_to_load = "~/Dropbox/Cerebro_development/pbmc_Seurat_dgCMatrix.crb"
)
```
